Current state and prospects of R-packages for the design of experiments

### 2022-06-15 paper/arxiv/arxiv.tex
done

\begin{abstract}
Re-running an experiment is generally costly and in some cases
impossible due to limited resources, so the design of an experiment
plays a critical role in increasing the quality of experimental data. In
this article we describe the current state of the R-packages for the
design of experiments through an exploratory data analysis of package
downloads, package metadata, and the comparison of characteristics with
other topics. We observe that experimental designs in practice appear to
be sufficiently manufactured by a small number of packages and the
development of experimental designs occur in silos. We discuss also the
interface designs of widely utilised R packages in the field of
experimental design and discuss its future prospects for the field.
\end{abstract}

@@

The critical role of data collection is well captured in the expression
``garbage in, garbage out'' -- in other words, if the collected data is
rubbish then no analysis, however complex it may be, can make something
out of it. A carefully crafted data collection scheme is therefore
critical to optimise the information from data. Of all the data
collection schemes, the field of experimental designs is specifically
devoted to planning an experiment largely based on the founding
principles by Fisher (1935) or an optimisation framework like those
described in Pukelsheim (2006). These experimental designs are often
constructed with the aid of a statistical software, such as R (R Core
Team 2021), so software usage can aid in understanding the practice of
experimental design. Given the large gain that can be made from
collecting higher quality experimental data, we benefit from
understanding the current state of the experimental designs in practice.

Methods for data collection can be dichotomised by the type of data
collected -- namely, experimental or observational -- or alternatively,
categorised as experimental design (including quasi-experimental design)
or survey design. This dichotomisation, to a great extent, is seen in
the \href{https://cran.r-project.org/web/views/}{Comprehensive R Archive
Network (CRAN) task views} (a volunteer maintained list of R-packages by
topic) where R-packages in experimental design are in
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view and R-packages in survey designs are in
\href{http://CRAN.R-project.org/view=OfficialStatistics}{\emph{OfficialStatistics}}
task view. The full list of available topics can be see in Table S1 in
the Supplementary Materials and also accessed via the \texttt{ctv}
package (Zeileis 2005). A subset of experimental designs are segregated
into
\href{http://CRAN.R-project.org/view=ClinicalTrials}{\emph{ClinicalTrials}}
task view, where the focus is on clinical trials with primary interest
in sample size calculations. This paper focuses on the packages in
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view, henceforth referred to as ``DoE packages''.

In
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view, there are 114 R packages for experimental design and analysis
of data from experiments. The sheer quantity and variation of
experimental designs in the R-packages are arguably unmatched with any
other programming languages, e.g.~in Python (Rossum 1995), only a
handful of packages that generate design of experiment exist (namely
\texttt{pyDOE}, \texttt{pyDOE2}, \texttt{dexpy}, \texttt{experimenter}
and \texttt{GPdoemd}) with limited type of designs. Thus, the study of
DoE packages, based on quantitative and qualitative data, can give us an
objective view to the state of the current experimental designs in
practice.

A utility of software can also be described by its design to facilitate
clear expression and interpretation of the desired experimental design.
Certain programming language design can hinder or discourage development
of a reliable program (Wasserman 1975). The immense popularity of
\texttt{tidyverse} (a collection of R-packages for various stages of
data analysis that place enormous emphasis on the interface design by
Wickham et al. 2019) is a testament to the impact an interface design
can have in practice.

The paper is organised as follows. Section \ref{data} briefly describes
the data source used for the analysis in Section \ref{eda}; Section
\ref{eda} presents some insights into the state of the current DoE
packages by exploratory data analysis of package download data, text
descriptions and comparisons with other CRAN task views; Section
\ref{design} discuss the interface designs of DoE packages, and we
conclude with a discussion in Section \ref{discussion} of future
prospects.

@@

The Comprehensive R Archive Network (CRAN) is a network of servers
located across the world that store mirrored versions of R and
R-packages. The most popular network is the RStudio mirror (the default
server for those that use the RStudio IDE). The RStudio mirror is also
the only server that provides a comprehensive daily download logs of R
and R-packages since October 2012. The summary data can be easily
accessed with the \texttt{cranlogs} package (Csárdi 2019). This paper
uses the data from the beginning of 2013 to end of 2021 (a total of 9
years) for the DoE packages.

@@

All CRAN packages have a title, description, package connections
(suggests, dependency and imports) and other meta-information in the
DESCRIPTION file. We use the text data from the title and description
(as accessed in 2022-05-23) in Section \ref{topics}.

@@

The list of packages in each CRAN task views (as of 2022-05-25) are used
to contrast the characteristics of DoE packages in Section \ref{silo}.

@@

All results presented are derived from exploratory data analysis of
observational data; consequently, all interpretations are somewhat
speculative and may not be indicative of the true state of the field of
experimental design. In particular, any analysis over time are
confounded by the fact that the nature of users and package management
have also changed over the years. It should be noted that some DoE
packages may have been archived or removed from the task view over the
years so any cross-sectional analysis presented may not reflect the set
of all DoE packages at that particular time period (although we assume
such incidences are low).

A subset of DoE packages is not primarily about design of experiments
but about the analysis of experimental data. A complete delineation of
these packages is difficult as many have some functions that can aid
decisions or constructions of experimental designs (and any
categorisation is prone to our subjective bias) so we opted not to
remove any DoE packages in the analysis.

\hypertarget{popular}{%
\subsection{Small, but diverse, set of packages are sufficient for most
designs}\label{popular}}

@@

\caption{Lorenz curve of the total download count for DoE packages in 2021. The red line corresponds to the line of perfect equality.}\label{fig:plot-lorenz}
\end{figure}

If we consider package downloads as a measure of ``wealth'', then we can
consider using the Gini index (Gini 1921) as a measure of download
inequality across packages. The ratio of the red region over the total
colored regions in Figure \ref{fig:plot-lorenz} corresponds to the Gini
index for 2021. A Gini index of 0\% indicates equality in downloads
across packages while a value of 100\% indicates maximal inequality (all
downloads are due to one package). In Figure \ref{fig:download-share},
we see that the distributions of the downloads each year have a heavy
right tail with the Gini index ranging from 32.5\% to 68.3\% across the
years 2013 to 2021, indicating that there is a high level of inequality
of downloads across packages, particularly with more pronounced
inequality in the last 5 years.

@@

\caption{The distribution of the number of downloads for each DoE package by year. Packages were removed in any year if it was released in that year or later so that each download count is for the full year. The label on the bottom of the plot shows the Gini index for downloads and the number of packages with full year of download count in the corresponding year. In the last 5 years, the Gini index is consistent above 60\% each year indicating that most downloads are due to a relatively small number of packages.}\label{fig:download-share}
\end{figure}

While in absolute terms the Gini index is high for
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}},
the inequality is not as severe as other CRAN task views as shown in
Figure \ref{fig:fig-gini-all-ctvs}. We can see in Figure
\ref{fig:fig-gini-all-ctvs} that the Gini index is generally increasing
for DoE packages (as is generally the case for other CRAN task views as
shown in Figure S1 in the Supplementary Materials).

@@

\caption{The points show the Gini index of the download counts by year for ExperimentalDesign CRAN task view with the color showing the number of packages. The grey line shows the distribution of the Gini index across years for all other CRAN task views. See Figure S2 in the Supplementary Material for the line graph for the Gini index across years by each CRAN task view.}\label{fig:fig-gini-all-ctvs}
\end{figure}

From Figures \ref{fig:download-share} and \ref{fig:fig-gini-all-ctvs},
we could draw some indicative states of the field of experimental
designs with possible counterfactual interpretations:

\begin{itemize}
\tightlist
\item
  The increase in the number of packages that are not highly downloaded
  may mean that \textbf{\emph{there are more packages to construct niche
  experimental designs}}. Some examples of these packages include
  \texttt{qtlDesign}, \texttt{PwrGSD} and \texttt{Crossover} made for
  QTL experiments, group sequential designs and crossover trials,
  respectively. These packages would naturally have a smaller number of
  potential users. Counterfactual to this, the increase could be due to
  other external factors, such as an increase in a number of skilled
  contributors, a change in CRAN policy or management to add packages
  (either to CRAN and/or task view), and/or that new packages are still
  yet to amass users. While there is an argument that low download
  counts are due to the low utility and/or quality of the packages,
  packages in CRAN task views are selected by expert maintainers; thus
  we can reasonably assume that any package listed in CRAN task view are
  of a decent utility and quality.
\item
  If the downloads are reflective of the experimental designs used in
  practice, \textbf{\emph{small set of packages appear to be sufficient
  for most to construct the full set of designs of experiments needed in
  practice}}. Packages of course evolve and the top downloaded packages
  have had regular updates that may have broadened its scope.
\item
  While small subset of DoE packages are most frequently used, none of
  these packages are dominant in the field (judging from comparisons
  with other CRAN task views). This suggests that there are
  \textbf{\emph{diverse approaches to designing experiments}}. This
  observation doesn't take into account other approaches to generating
  experimental designs, like the proprietary software, CycDesignN
  (Whittaker, Williams, and John 2022) and SAS (SAS Institute 1985),
  which may be widely used.
\end{itemize}

\hypertarget{ranking}{%
\subsection{There is a lack of adaptation of new or innovative
designs}\label{ranking}}

@@

\caption{The plot shows the rank of top 10 downloaded packages by year. }\label{fig:rank-over-time}
\end{figure}

Figure \ref{fig:established-packages} shows a moderate negative
correlation between the first release date and the (log of) total
download counts of DoE packages in any given year from 2013 to 2021.
This suggest that in general, a package released earlier are more likely
to be used today (possibly for legacy reasons or the general inertia for
adoption of new packages). We can also see in Figure
\ref{fig:established-packages} the most downloaded packages were
generally released in 2004 to 2010.

@@

{\centering \includegraphics{figures/established-packages-1}

@@

\caption{The above figure shows the total download (in log scale) of a package in the corresponding year against the first release date of the package. The blue line corresponds to the least squares fit of a simple linear regression model. The label in the right-hand upper corner shows the sample correlation coefficient between the first release date and the log, with base 10, of the total download count. The high-leverage point on the far left belongs to `conf.design`, authored by one of the earlier contributors to R.}\label{fig:established-packages}
\end{figure}

The lack of change in the top ranking packages are indicative that there
has been generally little need for new or innovative experimental design
needs for the mass. We do however also see that the the top downloaded
packages generally have more updates so it is possible that the packages
have improved or broadened the scope of its usage.

\hypertarget{topics}{%
\subsection{Computer experiments and optimal designs are of
interest}\label{topics}}

Figure \ref{fig:wordcloud-over-time} shows some common purposes of DoE
packages, based on bigrams in the package title and description. A
bigram is only shown as the frequency of single word was not insightful
and there were not many trigrams common across packages. Not
surprisingly, the bigram ``experimental design'' was the most common.
More interestingly, ``optimal design'' appeared across different
packages (indicated by the size of the word in Figure
\ref{fig:wordcloud-over-time}), and the bigrams ``latin hypercube'' and
``computer experiment'' used across a few packages that are downloaded
frequently (indicated by the color of the word in Figure
\ref{fig:wordcloud-over-time}). Although there exists a separate
\href{http://CRAN.R-project.org/view=ClinicalTrials}{\emph{ClinicalTrials}}
task view, the DoE packages clearly include some packages that are of
interest to clinical trials as shown by the size of the bigram
``clinical trial'' in Figure \ref{fig:wordcloud-over-time}.

@@

\caption{The above figure shows the word cloud of bigrams from the title and descriptions of the CRAN packages. The size shows how often the bigram appears across the DoE packages and the color are relative to the total download count in 2021 of the packages that contain the bigram.}\label{fig:wordcloud-over-time}
\end{figure}

@@

Figure \ref{fig:ctv-summ-plot} shows that the
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
have the lowest average number of contributors among all CRAN task
views. In addition, we can also see in Figure \ref{fig:ctv-summ-plot}
that the
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view has one of the least intra-connectivity (the percentage of
packages that make use of other packages within the same task view). The
full connection of DoE packages with other DoE packages is shown in
Figure \ref{fig:plot-doe-network}. These observations suggest that the
field of experimental design is one of the least collaborative field and
package development generally occur in silos.

@@

\caption{Package connections (depends, suggests and imports) within the DoE packages. DoE packages that do not depend, suggest or import another DoE package is not shown.}\label{fig:plot-doe-network}
\end{figure}

@@

In software design, there are two interface designs to consider: user
interface (UI) and application programming interface (API). UI is
concerned with the interaction of the software by the user, while the
API is concerned with how different programs interact and is
predominately of the interest to the developer. The UI design is an
abstraction to specifying the desired experimental design and its
choices enable how a user expresses the specification of the
experimental design.

In this section, we discuss these interface designs broadly, with
examples illustrated from the top downloaded packages (shown in Figure
\ref{fig:rank-over-time}). We exclude \texttt{ez} and
\texttt{DiceKriging} from the discussion as the former is predominately
visualisation of experimental data and latter is about the analysis of
computer experiments in addition to belonging to the same suite of
packages as \texttt{DiceDesign}.

\texttt{DoE.wrapper} (Grömping 2020)

\hypertarget{the-case-of-space-filling-and-optimal-designs}{%
\subsection{The case of space-filling and optimal
designs}\label{the-case-of-space-filling-and-optimal-designs}}

Computer experiments, as discussed in Section \ref{topics}, which
generally involve space filling designs, have been on the rise. The
exemplar for this type of designs are \texttt{lhs} and
\texttt{DiceDesign}. For the \texttt{lhs} package, functions like
\texttt{randomLHS()}, \texttt{optimumLHS()}, and \texttt{maximinLHS()},
require users to specify the sample size (\(n\)) and the number of
variables (\(p\)) then it generates a Latin hypercube sample, based on
different optimisation schemes (in this case, random, S optimal and
maxmin criteria, respectively; see package documentation for more
details). Similarly for the \texttt{DiceDesign}, there is a
comprehensive list of space-filling designs such as
\texttt{dmaxDesign()}, \texttt{lhsDesign()}, \texttt{straussDesign()}
and \texttt{wspDesign()} with input of \(n\) and \(p\) as before (among
extra parameters for some) that implement algorithms that either
maximise the entropy, produce a random Latin hypercube sample, based on
Strauss process and WSP algorithm, respectively. These designs all
output a \(n \times p\) matrix with values between 0 and 1.

@@

Factorial experiments offer a challenge in the construction and
allocation of the treatment factors. This is reflected in a number of
packages that specifically addresses this challenge. These include
\texttt{DoE.base} (Grömping 2018) for full factorial and (regular and
irregular) orthogonal array designs via \texttt{fac.design()} and
\texttt{oa.design()}, respectively, \texttt{FrF2} (Grömping 2014) for
fractional 2-level factorial designs using \texttt{FrF2()},
\texttt{FrF2Large()} or \texttt{pb()} Plankett-Burnman designs (Plackett
and Burman 1946), \texttt{conf.design} (Venables 2013) for symmetric
confounded factorial designs via \texttt{conf.design()}, and
\texttt{BHH2} (Barrios 2016) to generate full or fractional 2-level
factorial design matrix via \texttt{ffDesMatrix()}. All these designs
generally requires the user to input the number of factors and if the
design is allowed to vary in the number of levels, the input of the
number of levels for each factor. The construction of some of these
designs are catalogue-based, e.g.~\texttt{oa.design()} in
\texttt{DoE.base} and the entire package \texttt{FrF2.catlg128}
(Grömping 2022) is a list of fractional 2-level factorial designs.

@@

\hypertarget{the-case-of-menu-functions}{%
\subsection{The case of menu
functions}\label{the-case-of-menu-functions}}

@@

\hypertarget{acknowledgement}{%
\section{Acknowledgement}\label{acknowledgement}}

This paper uses the \texttt{targets} framework (Landau 2021) for
reproducibility, \texttt{knitr} (Xie 2015) and \texttt{rmarkdown} (Xie,
Allaire, and Grolemund 2018) for creating reproducible documents,
\texttt{ggplot2} (Wickham 2016), \texttt{ggraph} (Pedersen 2021),
\texttt{ggwordcloud} (Le Pennec and Slowikowski 2019) and
\texttt{colorspace} (Zeileis et al. 2020) for visualisation,
\texttt{kableExtra} (Zhu 2021) for customising the table in the
Supplementary Material and \texttt{tidyverse} (Wickham et al. 2019),
\texttt{tidytext} (Silge and Robinson 2016), \texttt{pluralize} (Rudis
and Embrey 2020), and \texttt{ineq} (Zeileis 2014) for data processing
and manipulation, and \texttt{cranlogs} (Csárdi 2019) and \texttt{ctv}
(Zeileis 2005) for extracting data. All code to reproduce this paper is
found at \url{https://github.com/emitanaka/paper-DoE-review}.

@@

@@

>>>>

\begin{abstract}
Re-running an experiment is generally costly and in some cases
impossible due to limited resources, so the design of an experiment
plays a critical role in increasing the quality of experimental data. In
this paper we describe the current state of the R-packages for the
design of experiments through an exploratory data analysis of package
downloads, package metadata, and the comparison of characteristics with
other topics. We observe that experimental designs in practice appear to
be sufficiently manufactured by a small number of packages and the
development of experimental designs often occur in silos. We discuss
also the interface designs of widely utilised R packages in the field of
experimental design and discuss its future prospects to advance the
field in practice.
\end{abstract}

@@

The critical role of data collection is well captured in the expression
``garbage in, garbage out'' -- in other words, if the collected data is
rubbish then no analysis, however complex it may be, can make something
out of it. A carefully crafted data collection scheme is therefore
critical to optimise the information from data. The field of
experimental designs is specifically devoted to planning the collection
of experimental data, largely based on the founding principles by Fisher
(1935) or an optimisation framework like those described in Pukelsheim
(2006). These experimental designs are often constructed with the aid of
a statistical software, such as R (R Core Team 2021), Python (Rossum
1995), SAS (SAS Institute 1985), and so on, so the usage of experimental
design software can inform us about some aspects of experimental designs
in practice.

Methods for data collection can be dichotomised by the type of data
collected -- namely, experimental or observational -- or alternatively,
categorised as experimental design (including quasi-experimental design)
or survey design. This dichotomisation, to a great extent, is seen in
the \href{https://cran.r-project.org/web/views/}{Comprehensive R Archive
Network (CRAN) task views} (a volunteer maintained list of R-packages by
topic) where R-packages for experimental design are in
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view and R-packages for survey designs are in
\href{http://CRAN.R-project.org/view=OfficialStatistics}{\emph{OfficialStatistics}}
task view. The full list of available topics can be see in Table S1 in
the Supplementary Materials. A subset of experimental designs are
segregated into
\href{http://CRAN.R-project.org/view=ClinicalTrials}{\emph{ClinicalTrials}}
task view, where the focus is on clinical trials with primary interest
in sample size calculations. This paper focuses on the packages in
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view, henceforth referred to as ``DoE packages''.

In
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view, there are 114 R packages for experimental design and analysis
of data from experiments. The sheer quantity and variation of
experimental designs in the R-packages are arguably unmatched with any
other programming languages, e.g.~in Python, only a handful of packages
that generate design of experiment exist (namely \texttt{pyDOE},
\texttt{pyDOE2}, \texttt{dexpy}, \texttt{experimenter} and
\texttt{GPdoemd}) with limited type of designs. Thus, the study of DoE
packages, based on quantitative and qualitative data, can give us an
objective view to the state of the current experimental designs in
practice.

A utility of software can also be described by its design to facilitate
clear expression and interpretation of the desired experimental design.
Certain programming language design can hinder or discourage development
of a reliable program (Wasserman 1975). The immense popularity of
\texttt{tidyverse} (a collection of R-packages for various stages of
data analysis that place enormous emphasis on the interface design by
Wickham et al. 2019) is a testament to the impact an interface design
can have in practice. The practice of experimental design could be
advanced by adopting similar interface design principles across DoE
packages.

The paper is organised as follows. Section \ref{data} briefly describes
the data source used for the analysis in Section \ref{eda}; Section
\ref{eda} presents some insights into the state of the current DoE
packages by the exploratory data analysis of package download data, text
descriptions and comparisons with other CRAN task views; Section
\ref{design} discusses the interface designs of widely used DoE
packages, and we conclude with a discussion in Section \ref{discussion}
of future prospects in the software development of experimental designs.

@@

The Comprehensive R Archive Network (CRAN) is a network of servers
located across the world that store mirrored versions of R and
R-packages. The most popular network is the RStudio mirror (the default
server for those that use the RStudio IDE). The RStudio mirror is also
the only server that provides a comprehensive daily download logs of R
and R-packages since October 2012. The summary data can be easily
accessed with the \texttt{cranlogs} package (Csárdi 2019). This paper
uses the data from the beginning of 2013 to end of 2021 (a total of 9
years) for the packages in the CRAN task views.

@@

All CRAN packages have a title, description, package connections
(suggests, depends and imports) and other meta-information in the
DESCRIPTION file. We use the text data from the title and description
(as accessed in 2022-05-23) in Section \ref{topics}.

@@

CRAN task views are volunteer maintained list of R-packages on CRAN
relevant to the corresponding topic. There are a total 37 CRAN task
views. Table S1 in the Supplementary Materials list the full available
topics from the \texttt{ctv} package (Zeileis 2005). The list of
packages in each CRAN task views (as of 2022-05-25) are used to contrast
the characteristics of DoE packages in Sections \ref{popular} and
\ref{silo}.

@@

In this section, we derive some conjecture based on the analysis of the
data described in Section \ref{data}. All results presented are from the
exploratory data analysis of observational data, consequently, all
interpretations are somewhat speculative and may not be indicative of
the true state of the field of experimental design. In particular, any
analysis over time are confounded by the fact that the nature of users
and package management have also changed over the years. It should be
noted that some DoE packages may have been archived or removed from the
task view over the years so any cross-sectional analysis presented may
not reflect the set of all DoE packages at that particular time period
(although we assume such incidences are low).

A subset of DoE packages is not primarily about design of experiments
but about the analysis of experimental data. A complete delineation of
these packages is difficult as there is almost always at least one
function that can aid decisions or constructions of experimental designs
(and any categorisation is prone to our subjective bias) so we opted not
to remove any DoE packages in the analysis.

\hypertarget{popular}{%
\subsection{Small, but diverse, set of packages are sufficient for most
experimental designs in practice}\label{popular}}

@@

\caption{Lorenz curve of the total download count for DoE packages in 2021. The red line corresponds to the line of perfect equality. The green region shows the area under the Lorenz curve and the red region shows the area of gap in equality.}\label{fig:plot-lorenz}
\end{figure}

If we consider package downloads as a measure of ``wealth'', then we can
consider using the Gini index (Gini 1921) as a measure of download
inequality across packages. The ratio of the red region over the total
colored regions in Figure \ref{fig:plot-lorenz} corresponds to the Gini
index for 2021. A Gini index of 0\% indicates equality in downloads
across packages while a value of 100\% indicates maximal inequality (all
downloads are due to one package). In Figure \ref{fig:download-share},
we see that the distributions of the package downloads each year have a
heavy right tail with the Gini index ranging from 42.6\% to 74.1\%
across the years 2013 to 2021, indicating that there is a high level of
inequality of package downloads, particularly with more pronounced
inequality in the last 6 years.

@@

\caption{The distribution of the number of downloads for each DoE package by year. Packages were removed in any year if it was released in that year or later so that each download count is for the full year. The label on the bottom of the plot shows the Gini index for downloads and the number of packages with full year of download count in the corresponding year. In the last 6 years, the Gini index is consistent above 60\% each year indicating that most downloads are due to a relatively small number of packages. }\label{fig:download-share}
\end{figure}

The increase in the number of packages that are not highly downloaded
may mean that \textbf{\emph{there are more packages to construct niche
experimental designs}}. Some examples of these packages include
\texttt{qtlDesign}, \texttt{PwrGSD} and \texttt{Crossover} made for QTL
experiments, group sequential designs and crossover trials,
respectively. These packages would naturally have a smaller number of
potential users. Counterfactual to this, the increase could be due to
other external factors, such as an increase in the number of skilled
developers (and thus more pacakge contributions), a change in CRAN
policy or management to add packages (either to CRAN and/or task view),
and/or that new packages are still yet to amass users. While there is an
argument that low download counts are due to the low utility and/or
quality of the packages, packages in CRAN task views are selected by
expert maintainers, thus we can reasonably assume that any package
listed in CRAN task view are of a decent utility and quality.

If the downloads are reflective of the experimental designs used in
practice, \textbf{\emph{small set of packages appear to be sufficient
for most to construct the full set of designs of experiments needed in
practice}}. Packages of course evolve and the top downloaded packages
have had regular updates that may have broadened its scope from its
previous releases.

While in absolute terms the Gini index is high for
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view (42.6\% to 74.1\%), the inequality is not as severe as other
CRAN task views as shown in Figure \ref{fig:fig-gini-all-ctvs}. We can
see in Figure \ref{fig:fig-gini-all-ctvs} that the Gini index is
generally increasing for DoE packages (as is generally the case for
other CRAN task views as shown in Figure S1 in the Supplementary
Materials) but most other CRAN task views have a Gini index of over
75\%. This suggests that other CRAN task views may have dominant
standards and in comparison to other topics, there are more
\textbf{\emph{diverse approaches to designing experiments}} and thus no
single DoE package is dominant. This observation however doesn't take
into account other approaches to generating experimental designs, like
the proprietary software, CycDesignN (Whittaker, Williams, and John
2022) and SAS (SAS Institute 1985), which may be widely used.

@@

\caption{The points show the Gini index of the download counts by year for ExperimentalDesign task view with the color showing the number of packages. The grey line shows the distribution of the Gini index across years for all other CRAN task views. See Figure S1 in the Supplementary Material for the line graph for the Gini index across years by each CRAN task view.}\label{fig:fig-gini-all-ctvs}
\end{figure}

\hypertarget{ranking}{%
\subsection{The field of experimental design is
slow-changing}\label{ranking}}

@@

\caption{The plot shows the rank of top 10 downloaded packages by year. Packages that do not appear in top 10 for at least two periods are omitted from the plot. Most packages are consistently in the top 10 for the period shown.}\label{fig:rank-over-time}
\end{figure}

Figure \ref{fig:release-date-vs-download} shows a moderate negative
correlation between the first release date and the (log of) total
download counts of DoE packages in any given year from 2013 to 2021.
This suggest that in general, a package released earlier are more likely
to be used today (possibly for legacy reasons or the general inertia to
adopt new packages). We can also see in Figure
\ref{fig:release-date-vs-download} the most downloaded packages were
generally released in 2004 to 2010.

@@

{\centering \includegraphics{figures/release-date-vs-download-1}

@@

\caption{The above figure shows the total download (in log scale) of a package in the corresponding year against the first release date of the package. The blue line corresponds to the least squares fit of a simple linear regression model. The label in the right-hand upper corner shows the sample correlation coefficient between the first release date and the log (with base 10) of the total download count. The high-leverage point on the far left belongs to `conf.design`, authored by one of the earlier contributors to R.}\label{fig:release-date-vs-download}
\end{figure}

The consistency in the top 10 ranking packages (Figure
\ref{fig:rank-over-time}) and the fact that most downloaded DoE packages
are first released more than 10 years ago (Figure
\ref{fig:release-date-vs-download}) indicates that either the existing
packages filled the needs of mass in practice or no new packages were
compelling for many to switch their practice. We do however also see
that the the top downloaded packages generally have more updates (see
Figure \ref{fig:release-date-vs-download}) so it is possible that the
packages have improved or broadened the scope of its usage.

\hypertarget{topics}{%
\subsection{Optimal designs are of interest}\label{topics}}

Figure \ref{fig:wordcloud-over-time} shows some common purposes of DoE
packages, based on bigrams in the package title and description. We show
only the bigrams as unigram was not insightful and there were not many
trigrams common across packages. In counting the bigrams, we processed
the text data as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  We standardised the words to lower case and removed pluralization.
\item
  Multiple mentions of the same bigram within a package was counted as
  one (e.g.~\texttt{AlgDesign} mentions ``experimental design'' four
  times in the title and the description but this is counted as one).
\item
  Bigrams that consist the stop words (provided in
  \texttt{tidytext::stop\_words} in addition to other words we deemed
  irrelevant, e.g.~``provide'', ``e.g.'', ``calculate'' and so on; full
  list is shown in the code provided in the link under Section
  \ref{pkgs}) are removed.
\end{enumerate}

Not surprisingly, the bigram ``experimental design'' was the most
common. More interestingly, ``optimal design'' and ``sequential design''
appeared across different packages (indicated by the size of the word in
Figure \ref{fig:wordcloud-over-time}), and the bigrams ``latin
hypercube'' and ``computer experiment'' are used across a few packages
that are downloaded frequently (indicated by the color of the word in
Figure \ref{fig:wordcloud-over-time}). Sequential design, Latin
hypercube sampling and computer experiments (which generally include
space filling designs like Latin hypercube sampling) are designs that
generally operate by optimising a user-selected criterion and can be
classified as optimal designs.

@@

\caption{The above figure shows the word cloud of bigrams from the title and descriptions of the DoE packages. The size shows how often the bigram appears across the DoE packages and the color are relative to the total download count in 2021 of the packages that contain the bigram.}\label{fig:wordcloud-over-time}
\end{figure}

Although there exists a separate
\href{http://CRAN.R-project.org/view=ClinicalTrials}{\emph{ClinicalTrials}}
task view, the DoE packages clearly include some packages that are of
interest to clinical trials as shown by the size of the bigram
``clinical trial'' (and related bigrams like ``dose finding'' and
``phase ii'') in Figure \ref{fig:wordcloud-over-time}.

@@

Figure \ref{fig:ctv-summ-plot} shows that the
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view has the lowest average number of contributors among all 37
CRAN task views. In addition, we can also see in Figure
\ref{fig:ctv-summ-plot} that the
\href{http://CRAN.R-project.org/view=ExperimentalDesign}{\emph{ExperimentalDesign}}
task view has one of the least intra-connectivity (the percentage of
packages that make use of other packages within the same task view). The
full connection of DoE packages with other DoE packages is shown in
Figure \ref{fig:plot-doe-network}. These observations suggest that the
\textbf{\emph{field of experimental design is one of the least
collaborative field}} and package development generally occur in silos.

@@

\caption{Package connections (depends, suggests and imports) within the DoE packages. The direction of the arrow shows the connection of packags where the package on the tail of the arrow is a depedency, suggestion or import for the package on the head of the arrow.  DoE packages that do not depend, suggest or import another DoE package is not shown.}\label{fig:plot-doe-network}
\end{figure}

@@

In software design, there are two interface designs to consider: user
interface (UI) and application programming interface (API). UI is
concerned with the interaction of the software by the user, while the
API is concerned with how different programs interact and is
predominately of the interest to the developer. The UI design is an
abstraction to specifying the desired experimental design and its
choices enable how a user expresses the specification of the
experimental design. The API design aids in other developers to leverage
existing systems.

In this section, we discuss these interface designs broadly, with
examples illustrated from the top downloaded packages (shown in Figure
\ref{fig:rank-over-time}). We exclude \texttt{ez} and
\texttt{DiceKriging} from the discussion as the former is predominately
visualisation of experimental data and the latter is about the analysis
of computer experiments in addition to belonging to the same suite of
packages as \texttt{DiceDesign} (Dupuy, Helbert, and Franco 2015).

Factorial experiments offer a challenge in the construction and
allocation of the treatment factors. This is reflected in a number of
packages that specifically addresses this challenge. These include
\texttt{DoE.base} (Grömping 2018) for full factorial and (regular and
irregular) orthogonal array designs via \texttt{fac.design()} and
\texttt{oa.design()}, respectively, \texttt{FrF2} (Grömping 2014) for
fractional 2-level factorial designs using \texttt{FrF2()},
\texttt{FrF2Large()} or \texttt{pb()} Plankett-Burnman designs (Plackett
and Burman 1946), \texttt{conf.design} (Venables 2013) for symmetric
confounded factorial designs via \texttt{conf.design()}, and
\texttt{BHH2} (Barrios 2016) to generate full or fractional 2-level
factorial design matrix via \texttt{ffDesMatrix()}. All these designs
generally requires the user to input the number of factors and if the
design is allowed to vary in the number of levels, the input of the
number of levels for each factor. The construction of some of these
designs are catalogue-based, e.g.~\texttt{oa.design()} in
\texttt{DoE.base}.

Saturated designs

\hypertarget{the-case-of-menu-designs}{%
\subsection{The case of menu designs}\label{the-case-of-menu-designs}}

Computer experiments, as discussed in Section \ref{topics}, which
generally involve space filling designs, have been on the rise. The
exemplar for this type of designs are \texttt{lhs} (Carnell 2022) and
\texttt{DiceDesign}. For the \texttt{lhs} package, functions like
\texttt{randomLHS()}, \texttt{optimumLHS()}, and \texttt{maximinLHS()},
require users to specify the sample size (\(n\)) and the number of
variables (\(p\)) then it generates a Latin hypercube sample, based on
different optimisation schemes (in this case, random, S optimal and
maxmin criteria, respectively; see package documentation for more
details). Similarly for the \texttt{DiceDesign}, there is a
comprehensive list of space-filling designs such as
\texttt{dmaxDesign()}, \texttt{lhsDesign()}, \texttt{straussDesign()}
and \texttt{wspDesign()} with input of \(n\) and \(p\) as before (among
extra parameters for some) that implement algorithms that either
maximise the entropy, produce a random Latin hypercube sample, based on
Strauss process and WSP algorithm, respectively. These designs all
output a \(n \times p\) matrix with values between 0 and 1.

@@

@@

@@

\hypertarget{pkgs}{%
\section{Acknowledgement}\label{pkgs}}

This paper uses \texttt{targets} (Landau 2021) and \texttt{renv} (Ushey
2022) for reproducibility, \texttt{knitr} (Xie 2015) and
\texttt{rmarkdown} (Xie, Allaire, and Grolemund 2018) for creating
reproducible documents, \texttt{ggplot2} (Wickham 2016), \texttt{ggraph}
(Pedersen 2021), \texttt{ggwordcloud} (Le Pennec and Slowikowski 2019)
and \texttt{colorspace} (Zeileis et al. 2020) for visualisation,
\texttt{kableExtra} (Zhu 2021) for customising the table in the
Supplementary Material and \texttt{tidyverse} (Wickham et al. 2019),
\texttt{tidytext} (Silge and Robinson 2016), \texttt{pluralize} (Rudis
and Embrey 2020), and \texttt{ineq} (Zeileis 2014) for data processing
and manipulation, and \texttt{cranlogs} (Csárdi 2019) and \texttt{ctv}
(Zeileis 2005) for extracting data. All code to reproduce this paper is
found at \url{https://github.com/emitanaka/paper-DoE-review}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-lhs}{}}%
Carnell, Rob. 2022. \emph{Lhs: Latin Hypercube Samples}.
\url{https://CRAN.R-project.org/package=lhs}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-DiceDesign}{}}%
Dupuy, Delphine, Céline Helbert, and Jessica Franco. 2015.
{``{DiceDesign} and {DiceEval}: Two {R} Packages for Design and Analysis
of Computer Experiments.''} \emph{Journal of Statistical Software} 65
(11): 1--38. \url{https://www.jstatsoft.org/v65/i11/}.
